Perennial is a framework for reasoning about crash safety and concurrency that
we developed in order to verify GoTxn. The main component of Perennial is a
program logic based on concurrent separation logic, with extensions for
reasoning about crash and recovery behaviors.

\paragraph{Who is this chapter for?}
This chapter describes Perennial for an audience with some programming languages
or verification background, but not necessarily experience with concurrency
specifically. The high-level ideas are meant to be broadly accessible even if
some background on Iris is needed to appreciate the details. To that end the
presentation is somewhat simplified, with details and side conditions omitted
from the logical rules. A more systems-oriented reader can safely skip this
chapter and the next (\cref{ch:crash-logatom}) and still understand the later
chapters.

\input{perennial/01-iris}
\input{perennial/02-logic}

\section{Limitations}
\label{sec:perennial:limitations}

Perennial has some limitations. The logic can prove \emph{safety} properties
(``nothing bad happens'') but not \emph{liveness} (``something good eventually
happens''), a limitation inherited from Iris and in fact due to deeper
theoretical reasons related to step indexing --- Transfinite Iris is an approach
to solve this issue that discusses the reasons well~\cite{spies:transfinite}.
Due to different technical reasons related to step indexing, it is not currently
possible to put weakest precondition assertions inside crash conditions; such a
feature is useful for reasoning about procedures stored on disk, something that
shows up in practice with logical logging~\cite{mohan:aries}.

The current version of Perennial does not support \emph{recovery helping} from
the original version~\cite{chajed:perennial} (although it has many new
features), a reasoning principle where in a simulation proof recovery logically
completes an operation from before the crash. As described here, operations must
be simulated during the crash. It also does not implement Iris's support for
prophecy variables~\cite{jung:prophecy}, though we believe this is doable --- an
interesting extension to work out would be prophecies that predict a crash, or
prophesize actions during recovery.

\section{Takeaways beyond formal verification}

\paragraph{Crash-aware locks.} One of the benefits of concurrent separation
logic that is useful even without verification is that it crystallizes the idea
of a ``lock invariant'' as a way to think about the \emph{purpose} of locks.
Rust's \cc{sync::Mutex} API implements a basic version of the lock invariant
idea by allowing the programmer to at least express what data is protected by
the lock, but the idea of a separation logic invariant is more general (for
example, locks can protect just a part of a data structure). Perennial's
crash-aware lock specification formalizes an idea that is not well-known, of how
locking lock invariants interact with durable state: such a locks should have
both the usual lock invariant and a weaker crash invariant.

\paragraph{Invariants for crash reasoning.} Perennial has a rule where
allocating an invariant gives ``credit'' for it on crash. This idea formalizes
an intuition about how ownership flows between threads and on crash. A system
called RECIPE has an insight along these lines, observing that a lock-free data
structure should work on persistent memory~\cite{lee:recipe}. Intuitively this
is true because the code already maintains an invariant at all intermediate
steps due to other threads, and thus the same invariant should hold of the
persistent memory on crash, but some more conditions are required for the
intuition to hold up. RECIPE has no proof of correctness, and a mechanized proof
would be challenging, but the ideas in this thesis could be used to think about
its correctness and the required conditions with greater precision.
