On top of the Perennial logic, we developed a specification pattern to organize
proofs of libraries and compose them. Of particular interest is a pattern for
\emph{logically-atomic crash specifications}, which capture that a set of
methods in a library appear atomic with respect to both other threads and on
crash. In order to illustrate how all of these techniques work together, this
section focuses on the specification and proof intuition for the circular
buffer, a library used in GoTxn that has interesting crash safety and
concurrency behavior. One feature of this specification style, the idea of an
\emph{exchanger}, only appears in the journaling code which is much higher in
the software stack for GoTxn, so \cref{sec:perennial:exchanging} does discuss a
more sophisticated specification. The circular buffer is relatively independent
of other details in GoTxn, since it operates directly on top of the disk, but
journaling is covered in more detail within the GoTxn chapter, in
\cref{sec:txn:lifting}.

This chapter builds upon the Perennial background from \cref{ch:perennial}, and
the note about audience from that chapter's introduction is relevant here as
well --- a reader interested in the ideas in GoTxn, DaisyNFS, or Goose can
safely skip to the relevant chapters.

\input{crash-logatom/01-circ.tex}
\input{crash-logatom/02-spec-style.tex}

\section{Take-aways beyond formal verification}

\paragraph{Physical versus logical logging.} The GoTxn log uses physical
logging, where log entries are 4KB disk writes. An alternative is to use logical
logging, where entries instead correspond to higher-level operations like
appending to a file. From a correctness (and especially crash safety)
perspective, what makes logical logging challenging is that it appears to break
modularity: the concept of a file now shows up in the write-ahead log.

The techniques in Iris and this chapter give a way to think about logical
logging modularly where the log stores \emph{procedures}. The insight is that
the calling code passes a function to the write-ahead log that interprets the
logical log entry into block operations. This makes logical logging a
higher-order interface. Note that in practice the code may not be organized this
way; we can use \emph{defunctionalization} to specialize the logging code to
only the procedures used by the caller, especially for the common use case of a
tightly integrated file system and journal. The write-ahead log will rely on
some specification (including a crash condition) for the function that
interprets its entries. In this simplest case it might require idempotent
operations; a more sophisticated logging design like ARIES~\cite{mohan:aries}
can handle non-idempotent operations using a combination of redo logging (like
GoTxn) and undo logging.
