GoTxn is a transaction system we implemented and verified with the goal of
making crash safety and concurrency simple for a storage system implemented on
top. Transactions appear to run atomically both on crash and to other threads.
The GoTxn specification formalizes this property, which we use in DaisyNFS to
enable sequential reasoning for a concurrent file system (described in
\cref{ch:daisy-nfs}).

This chapter describes GoTxn's implementation, top-level specification, and
several aspects of the proof. The implementation is composed of a software stack
with several layers; each layer implements a new interface on top of one
below. The proof follows the same structure, with several new specifications for
intermediate layers. Of particular note are the
transaction-refinement specification for the transaction system, the lifting
specification for the journaling layer, and the abstraction for the write-ahead
log layer.

While as part of this thesis we only used GoTxn to implement and verify a file
system, the system and its specification are generic for any
storage system implemented on top, as long as its operations are implemented
using transactions.

GoTxn faces three key challenges in its specification, design, and proof.
First, GoTxn's specification is stated in terms of programs using the
transaction system, and not all programs will observe atomic transactions ---
for example, accessing global variables or the network in the middle of a
transaction is not atomic. To state a provable specification for GoTxn we needed
a formalization of \emph{safe transactions}, described in \cref{sec:txn:transaction-refinement},
which obey some restrictions that the specification assumes the caller follows.

The second challenge is supporting an in-memory allocator, which is necessary
for the file system to get good performance but appears to violate GoTxn's rule
that transactions do not access shared memory since allocating and
freeing affect other transactions immediately, not at commit time.
To address this challenge we include an
allocator in the GoTxn API and proof. A non-deterministic,
\emph{under-specification} of the allocator makes its behavior serializable, and
\cref{ch:daisy-nfs} demonstrates that this allocator is usable by using it in
the DaisyNFS verified file system.

Finally, this specification is related to a concrete implementation, and thus
requires a proof using a program logic. The proof manages the complexity of
GoTxn's implementation by combining a \emph{lifting-based specification} that
captures the journaling layer's crash atomicity (\cref{sec:txn:lifting}) with a
simulation proof that captures how two-phase locking's concurrency control works
(\cref{sec:txn:refinement}).

\input{go-txn/01-api}
\input{go-txn/02-spec}
\input{go-txn/03-system}
\input{go-txn/04-overview}
\input{go-txn/06-proof}

\section{Limitations}
\label{sec:txn:limitations}

GoTxn has some limitations. The transaction-refinement proof currently only support
\emph{synchronous} \cc{Commit}, which makes writes durable when it returns. The
code also implements an asynchronous version of \cc{Commit} that makes the
writes atomically visible to other threads, but durable only at a later point.
It would be interesting to extend the transaction-refinement specification and proof
to cover asynchrony. The two-phase locking implementation only supports write
locks and not reader-writer locking. At the bottom layer, GoTxn's disk interface
assumes a synchronous disk interface --- it would be interesting to relax this
to reason about asynchronous writes, which model how disks can lose recent
writes on power failure due to internal buffering.

\input{go-txn/99-concl}
