This chapter is about Goose, which solves a practical
problem of reasoning about efficient code. To use Goose, a developer writes
code in Go, uses an automatic translator to convert the code to a model in the Coq proof
assistant, then carries out the proof on top of the model in the Perennial logic. Goose encompasses
the entire process: it includes the translation tool itself, the way it models Go
code, and finally the reasoning principles for proving properties of translated
Go. (We will also use ``Goose'' in some places to refer to the subset of Go
supported by the translation tool.)

This chapter is intentionally fairly independent of the rest of the thesis for a
reader interested in verifying Go code but not crash safety or the specifics of
the GoTxn and DaisyNFS proofs. Crashes are modeled simply by stopping
execution and wiping out all of the state except for the disk, which does not
relate to the specifics of Go.

\input{goose/01-goals.tex}
\input{goose/02-related.tex}
\input{goose/03-overview.tex}
\input{goose/04-lang.tex}
\input{goose/06-reasoning.tex}
\input{goose/07-testing.tex}

\section{Limitations}%
\label{sec:goose:limitations}

Notably missing in Goose but prominent in Go is support for interfaces
and channels. We believe both are easy enough to support, but interfaces
were not necessary for our implementation, and rather than channels the code
verified in this thesis
use mutexes and condition variables for more low-level control over
synchronization.

Control flow is slightly tricky since a Go function is translated
to a single GooseLang expression that should evaluate to the function's
return value. Goose supports many specific patterns, especially common
cases like early returns inside \cc{if} statements and loops with
\cc{break} and \cc{continue}, but more complex control flow ---
particularly returning from within a loop --- is not supported. It would be
easiest to express general control flow in continuation-passing style (in which
every GooseLang takes a continuation, and calling this continuation corresponds
to returning from the function in Go), but this
would complicate every specification and the translation of function calls.

Goose do not support Go's \cc{defer} statement. It would be nice to support some
common and simple patterns, particularly for unlocking, by translating
\cc{defer} statically. The behavior of Go's \cc{defer} statement in general is
to push the deferred function to a stack of calls associated with the current function that are executed in reverse order at return
time. GooseLang does not have a first-class notion of a Go function to associate
the stack of deferred functions with, nor the concept of returning
from a function. However, it would be useful to simple static uses of \cc{defer}
at the top-level of a function.

Named return values are recommended to document return parameters, and sometimes
simplify and clarify the body of a function.\footnote{See the description in
\href{https://go.dev/doc/effective_go\#named-results}{Effective Go}.} However, in
general they are quite subtle, due to interaction with \cc{defer} statements and
concurrency~\cite{chabbi:golang-races}. One source of difficulty is that the
return values are treated like local variables declared at the top of the
function, and it is easy to accidentally have races on these variables if they
are accessed concurrently.

Goose does not support mutual recursion between Go functions, and
additionally requires the translation to be in the right order so
definitions appear before they are used. Files in a package are especially prone
to this issue, since they are processed in alphabetical order; we sometimes name
files something like ``0constants.go'' to make the order works out correctly.
The subtlety here is that
definition management in Go, as in most imperative languages,
conceptually treats all top-level definitions as simultaneous, whereas
Coq processes definitions sequentially. Using Coq definition management
to model Go definition management imposes a limitation compared to Go,
but is much simpler to work with compared to modeling a Go package as a
set of mutually recursive definitions. Reasoning about code written in such a
model would require setting up specification for all the definitions, then
proving them in a recursive way, all while ensuring that no specification is
used before it is proven.

GooseLang does have one extant bug related to evaluation contexts. The contexts
$e \app E$ and $E \app v$ define a right-to-left evaluation order for functions,
which is the opposite of Go. We haven't yet fixed this, either by adjusting the
GooseLang semantics or changing the translation to emit code that explicitly
evaluates all the arguments in the correct order before calling the function.

Map iteration in Go happens in a non-deterministic order.\footnote{In fact the
runtime randomizes the starting position of the iteration, to avoid callers
accidentally relying on any particular behavior. See
\href{https://github.com/golang/go/blob/c379c3d58d5482f4c8fe97466a99ce70e630ad44/src/runtime/map.go\#L844-L850}%
{\cc{mapiterinit} from src/runtime/map.go}.} Thus, strictly speaking, the model
of map iteration described in \cref{sec:goose:maps} given by \cc{MapIter} should
shuffle the elements of the map before iterating over it, in order to model the
non-determinism of the implementation. Goose does not (currently) do this,
simply because the shuffle would be hard to implement in GooseLang. However, the
specification for map iteration does not expose an iteration order and would
apply unchanged to this more non-deterministic model. All proofs go through a
common iteration specification, so our proofs should remain unchanged if
\cc{MapIter} started modeling non-deterministic ordering.

\section{Implementation}%
\label{sec:goose:impl}

Goose is implemented in Go. It takes advantage of a family of packages in under
\cc{go} in the Go standard library, such as \cc{go/ast} and \cc{go/types}, to
parse and type-check Go code; the \cc{golang.org/x/tools/go/packages} package
even makes it possible to work with Go packages and modules. These greatly
simplify implementing Goose and also make it more trustworthy, since at least
the source code is parsed in exactly the same way as the compiler (though we can
still introduce bugs in interpreting what the AST means).
The translator has an intermediate representation of the Coq source for
GooseLang, and splits translation into generating the Go structs for this
representation and subsequently printing this representation as a Coq file.

The translator makes as much of an attempt as possible to identify and report
errors, identifying where a feature is unsupported. Each function is translated
almost independently, allowing translation to move on to the next function and
report a whole batch of errors. Some issues are not checked by the translation,
such as the topological order on definitions mentioned in limitations above (in
section \cref{sec:goose:limitations}); in these cases the resulting Coq code
does not compile, which preserves soundness but is a slightly worse user
experience.

GooseLang has a notion of ``external'' operations and state, which is generally
instantiated with a disk. The translator checks that the source package imports
only one of the allowed external imports, including through its transitive
dependencies. It also supports code that uses no external operations,
translating it to a form that is itself parameterized, allowing the result to be
used as part of any other code.

\section{Conclusion}

Goose is an approach for verifying Go code. We define GooseLang as a model of Go
and automatically translate a subset of Go to this language. GooseLang comes
with a number of reasoning principles for handling features of Go. The benefit
of this approach is the ability to write high-performance code in a
productive language, with convenient reasoning while verifying
that code. Several aspects of the design contribute to making the approach
sound, ranging from the subset of Go supported, to the design of GooseLang, and
the use of standard Go tools for analyzing the source code.

Our main use case for Goose for this thesis was to verify GoTxn, but
the tool and approach are more generally applicable, even without concurrency or
crash-safety reasoning. These ideas could also be productively applied to
languages other than Go --- I am personally excited about the prospect of having
a version for Rust.
