\section{Programming with GoTxn}
\label{s:gotxn-api}

\begin{figure}
\begin{minted}{go}
type Addr struct {
  Blkno  uint64
  Offset uint64
}

// starting and stopping a transaction
func Begin() *Txn
func (tx *Txn) Abort()
func (tx *Txn) Commit()

// operations within a transaction
func (tx *Txn) Read(a Addr, sz uint64) []byte
func (tx *Txn) ReadBit(a Addr) bool
func (tx *Txn) Write(a Addr, d []byte)
func (tx *Txn) WriteBit(a Addr, d bool)

// allocator API
func NewAllocator(max uint64) *Allocator
func (a *Allocator) Alloc() uint64
func (a *Allocator) Free(n uint64)
\end{minted}
  \caption{The API for the transaction system and allocator. Reads and writes
    between \cc{Begin} and \cc{Commit} appears to execute atomically on disk and
    for other threads, while \cc{Abort} guarantees the transaction has no
    effect. The allocator's \cc{Alloc} and \cc{Free} operations are safe to call
    concurrently.}
\label{fig:txn-api}
\end{figure}

The transaction system gives the caller a way to read and write on-disk
objects (which can be smaller than the 4KB block size of the underlying disk),
in such a way that a whole sequence of reads and writes appears atomic. The
programming interface is listed in \cref{fig:txn-api}. In practice this makes
it easier to implement a correct storage system by handling the challenges of crash safety
and concurrency, leaving only correct implementing the data structures and logic
of the system using disk objects.

To use GoTxn, a storage system follows a similar pattern for each operation of
wrapping all of the code for that operation in a single transaction. Operations
start by creating an empty transaction with \cc{Begin()}.
The entire
transaction appears to execute atomically when the operation finishes it
with \cc{Commit}, or the transaction is discarded with no effect on
\cc{Abort}. Reads and writes operate on addresses which specify a
position by giving a block number and an offset in bits (always less
than $4096 \cdot 8$, the number of bits in a block). The \cc{Read}
method requires an explicit size argument while the size of a
\cc{Write} is implicit in the size of the \cc{data} slice. We separate
out the bit-sized operations to \cc{ReadBit} and \cc{WriteBit} (rather
than using a single-element byte slice) to simplify the specification.

\Cref{fig:txn-api} also includes an API for allocation alongside the
transaction API.\@ \Cref{ch:daisy-nfs} will explain in more detail why this
API must be part of the transaction system, since it is relevant to the
file-system design. \tej{should not forward reference DaisyNFS, explain this
somewhere in this chapter} The allocator should be considered part of the transaction
system API insofar as its operations are allowed within a transaction.
In contrast using other shared memory within a transaction is not permitted since it
would compromise the atomicity guarantees --- the transaction system acquires
locks to make reads and writes seem atomic, but it doesn't have any locking
discipline for other state.

%% The state of the transaction system (the transactional disk transactions
%% manipulate) looks much like a flat array of bytes.
%% However, the caller cannot
%% read and write arbitrary regions of this array due to restrictions in the
%% gojournal code and proof. all reads and writes must be within a single 4kb block
%% on disk, and of a power-of-two number of bytes or a single bit.

% In practice the file system uses three kinds of objects: full blocks are used
% for data (both for directories and data files), bit objects comprise the inode
% and block allocators, and 128-byte objects are used to represent inodes. The
% file-system statically allocates regions for the inodes, allocator bitmaps,
% and data blocks, so that object sizes never change.

As described next in \cref{s:gotxn:impl}, GoTxn is implemented using two-phase
locking. As a result, a transaction acquires a per-address lock on every address
it reads or writes along the way. Acquiring multiple locks during a transaction
creates the possibility for deadlocks, if two threads acquire a pair of locks in
the opposite order, and the specification does not forbid
deadlocks.\footnote{Liveness reasoning in general is quite challenging when
combined with concurrency. However, it would be interesting to specify and
verify deadlock freedom without verifying full liveness.} The two-phase locking
implementation does not implement a specific lock acquisition order, leaving it
to the calling code to avoid deadlock --- for example, in DaisyNFS the
implementation of \cc{RENAME} makes sure to lock the smaller inode number first
(by reading from it) if the rename is between different directories.
