\section{Implementing GoTxn}
\label{sec:txn:impl}

The transaction system is structured into several layers, as shown in
\cref{fig:gotxn-layers}. At a high level, three of the abstractions are useful
to understand the overall structure: the \scc{wal}, the \scc{jrnl}, and finally
the top-level \scc{txn}. The first useful abstraction is the
write-ahead log, which behaves like a disk with an atomic multiwrite operation.
Reads and writes still operate on 4KB blocks, but a multiwrite appears to update
multiple disk blocks simultaneously even if the system crashes. Next, the
\scc{jrnl} layer implements \emph{journaling}, persisting a whole operation with
reads and writes to disk atomically. Concurrent operations must access disjoint
sets of addresses for safety; concurrency control is left to the caller in this layer.
Operations can manipulate objects smaller than a block (``sub-block'' objects),
which improves concurrency by making more operations disjoint. Support for
sub-block objects is implemented in the \scc{obj} layer between the \scc{wal}
and \scc{jrnl}. Finally, the
\scc{txn} layer exports the complete transactional interface in
\cref{fig:txn-api}. This layer implements automatic concurrency control so that
the caller can freely read and write any addresses.

\begin{figure}[htb]
  \centering
  \small
  \begin{tabular}{ll}
    \toprule
    \textbf{Layer} & \textbf{Description} \\
    \midrule
    \scc{txn} & Transactions \\
                   & implements concurrency control using two-phase locking \\
    \scc{jrnl} & Journaling \\
                   & implements in-memory buffering \\
    \scc{obj} & Sub-block objects \\
                   & implements reads and atomic writes within a block \\
    \scc{wal} & Atomic whole-block multiwrites \\
                   & implements whole-block write-ahead logging \\
    \scc{circ} & On-disk queue with atomic append \\
                   & implements a durable circular log \\
    \midrule
  \end{tabular}
  \caption{The layers in the GoTxn implementation.}
  \label{fig:gotxn-layers}
\end{figure}

The write-ahead log is implemented by organizing the disk into a small,
fixed-size circular buffer and a remaining data region. Data is first atomically
\emph{logged} to the circular buffer and then eventually \emph{installed}
to the data region, to free space in the circular buffer. Reads first go through
the circular buffer (which is cached for efficiency) and then access the data
region. The circular buffer's implementation is described in greater detail in
\cref{sec:circ-api}.

The object system maintains a list of buffers of data read or written by each journal operation.
Reads first check the log (which is always cached in memory) since
they must observe committed operations. To commit, the object
layer gathers all the dirty buffers and submits them as a multiwrite to the
write-ahead log. To allow reading and writing objects that are smaller than a
block, the object layer assembles these into block writes by doing a
read-modify-write sequence.
%if a block isn't completely overwritten within an operation.

Because disk writes are slow, for good performance the journal executes many
tasks in parallel. Committing new journal operations in memory, logging operations
from memory to disk, waiting for operations to be made durable, and
installing logged writes all happen concurrently.  Concurrency ensures that
in-memory operations
need not wait for any in-flight disk reads or writes, and that many
disk reads and writes can happen at the same time.  Finally, to reduce the
number of disk writes, the write-ahead log implements two optimizations.
Multiwrites are combined and written
together (``group commit''), and if they update the same disk
block multiple times, only the most recent update of that disk block is
written to the log (``absorption''). Concurrency makes these optimizations
useful even for synchronous operations, which can be committed together and
absorbed if they are issued concurrently.

The \scc{obj} or object layer implements sub-block access on top of the write-ahead
log's block-level multiwrites. Objects accessed by an operation must be locked,
so supporting fine-grained access is necessary to allow operations to run
concurrently even if they happen to access the same disk block. For example, a
file system might pack inodes into a block, and locking an inode should not
prevent concurrent operations for other inodes in the same block. The
object-layer implementation is able to execute reads and writes during an
operation without any additional locks, but something more is needed to commit.
Imagine a situation where between reading some disk block and writing it an
unrelated object was modified in the same block; committing the modified block
would overwrite the concurrent modification, losing data. The code addresses
this with a global commit lock that prevents concurrent modifications while
reading the blocks to be written.

The \scc{jrnl} layer implements a journaling system which gives the caller a
useful abstraction over the disk that makes it easy to update the disk in a
crash atomic way but which requires that the caller implement appropriate
concurrency control. The \scc{txn} layer automatically implements the required
concurrency control; this isn't much code beyond the journal, but it does
dramatically change the specification since the caller sees any sequence of
reads and writes as atomic both with respect to threads and crashes (also known
as \emph{serializability}). The \scc{txn} layer ensures transactions don't
conflict using two-phase locking. This algorithm acquires a lock on each address
the first time it is used in a transaction. Writes are buffered locally until
commit time, at which point they are written atomically using the journaling
system. Finally all the locks are released, exposing the transaction's effects
to subsequent transactions.
