\section{Conclusion}
\label{sec:txn:concl}

GoTxn is a concurrent, crash-safe transaction system with a machine-checked
proof. The caller wraps a sequence of reads and writes in a transaction, which
the transaction system promises to make atomic. GoTxn has a relatively
sophisticated implementation. The write-ahead log supports buffering new writes,
logging for durability, and installation concurrently since the logger and
installer threads access disk without locks. The object layer implements
safe concurrent access to objects within the same block. Two-phase locking
ensures transactions do not conflict, guaranteeing isolation.

We formalized the system's atomicity guarantee in the form of a \emph{program
refinement} specification, which says that when the caller issues an arbitrary
sequence of transactions, they appear to execute atomically, even if
transactions are issued concurrently and if the system crashes. A key challenge
in this specification is formalizing ``safe'' transactions which GoTxn can make
atomic, forbidding arbitrary access to shared memory that would fall outside the
two-phase locking discipline. To make GoTxn practical with this restriction it
includes an in-memory allocator with a specification that uses non-determinism
to fit the allocator into GoTxn's atomicity specification.

GoTxn is verified against its program refinement specification using Perennial.
The proof is modular, using Perennial's logically atomic crash specifications to
verify each layer separately and compose the proofs together. Of particular note
in the proof is the \emph{lifting-based specification} specification of the
journaling layer that captures what concurrent transactions do.
